---
title: "BatchVaria Quickstart: Evaluating Batch Correction Performance"
author:
- name: Nicholas Moir
  affiliation: University of Edinburgh
  email: nicholas.moir@ed.ac.uk
package: BatchVaria
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{BatchVaria Quickstart: Evaluating Batch Correction Performance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, message=FALSE, warning=FALSE}
library(BatchVaria)
library(SummarizedExperiment)
```

# Overview

High-throughput transcriptomic datasets frequently contain unwanted technical variation arising from batch effects. Numerous batch correction methods exist, but selecting an appropriate method is challenging because correction can remove biological signal or distort sample relationships. BatchVaria provides a framework for quantifying variance components, tracking variance changes after correction, and evaluating correction performance using multiple metrics, enabling evidence-based selection of batch correction strategies.
<br>
<br>
While tools such as ComBat, limma, and BatchQC provide batch correction and diagnostic visualisations, BatchVaria focuses specifically on quantitative evaluation of correction performance across multiple assays and correction strategies within a unified framework. BatchVaria unifies variance decomposition, tracking, and comparison into a single coherent workflow. 
---

# Installation

```{r installation}
# Install from Bioconductor (devel)
# BiocManager::install("BatchVaria")
```

# Example Dataset

We begin with a synthetic dataset designed to reflect realistic batch correction scenarios. The example dataset contains simulated gene expression data with both biological group effects and batch effects. Several pre-generated corrected assays are included to illustrate different correction outcomes, including over-correction, under-correction and noise distortion. A ComBat correction is also performed as part of the workflow.


```{r runCorrection}
## exampleBatchVaria() draws random data; seed for a reproducible vignette
set.seed(1)
bv <- exampleBatchVaria()

bv <- runCorrection(bv, method = "combat", batch = "batch", assayName = "raw")

bv
```

```{r display_assays}
assayNames(bv)
```

There are now 5 assays contained within bv, the BatchVariaData class instance. The "raw" assay is uncorrected data, "raw_combat" is ComBat corrected, "raw_center" is mean-centred data, "raw_scale" is scaled data and "raw_noise" is a distorted correction.

---

# Variance Profiling 

We explicitly quantify the contribution of known covariates.

```{r profile}
bv <- profileVariance(bv, formula = ~ batch + group, assays = c("raw", "raw_combat", "raw_center", "raw_scale", "raw_noise"), methods = c("variancePartition"))
```
This step fits the specified model to each assay and decomposes variance into batch (unwanted), group (biological) and residual (unexplained). Results are stored within an internal variance ledger.

---

# Variance Summary Tables

A core output of BatchVaria is a pair of summary tables.

```{r table}
vt <- varianceTable(bv, assays = c("raw", "raw_combat", "raw_center", "raw_scale", "raw_noise"))
```
---

# Variance Composition Table

This table shows the percentage of variance explained by each component for each assay.

```{r percent}
vt$percent
```
---

# Variance Change Relative to Raw

This table shows how each correction changed variance relative to the raw data.

```{r delta}
varianceTable(
    bv,
    assays = c("raw", "raw_center", "raw_scale", "raw_noise", "raw_combat"),
    formatDelta = TRUE
)$delta
```
---

# Interpretation

The variance summary tables generated by BatchVaria provide a straightforward framework for interpreting the effects of batch correction on expression data. In general, an effective correction method should reduce the proportion of variance attributable to unwanted technical factors such as batch, while preserving or enhancing variance associated with biological variables of interest. At the same time, large increases in residual variance may indicate that the correction procedure has introduced noise or removed meaningful structure from the data.

In this example, mean-centring and scaling do not change the variance composition relative to the raw data, indicating that these transformations do not address batch effects in this dataset. The noisy correction reduces batch-associated variance, but it also reduces the variance attributed to the biological group variable and substantially increases residual variance, suggesting overcorrection and loss of biological signal. In contrast, the ComBat-corrected assay shows a strong reduction in batch variance while maintaining or slightly increasing group-associated variance, indicating successful removal of unwanted technical variation while preserving biological signal. Although residual variance increases slightly after correction, this is expected when structured technical variation is removed from the data.

Together, these results demonstrate how variance decomposition across multiple corrected assays provides a clear and quantitative framework for comparing batch correction strategies and identifying methods that best balance unwanted variation removal with biological signal preservation.

# Summary Figure
```{r fig.width=10, fig.height=7, out.width="100%"}
plotVarianceComposition(bv, method = "variancePartition")
```

# Why This Matters

Batch correction is frequently applied as a routine preprocessing step, yet it is often performed without systematic or quantitative validation. This can lead to suboptimal outcomes, where overcorrection removes meaningful biological signal or undercorrection leaves residual technical artefacts that confound downstream analyses. BatchVaria addresses this gap by providing a transparent and reproducible framework for evaluating correction strategies. By enabling direct comparison of variance components across multiple corrected assays, the package supports quantitative validation of correction outcomes and facilitates more robust, interpretable analytical pipelines.

---

# BatchVaria Workflow Summary

The typical BatchVaria workflow is:

1. Load expression data into BatchVariaData
2. Apply batch correction methods 
3. Profile variance using profileVariance()
4. Summarise variance using varianceTable()
5. Compare corrections using delta variance
6. Visualise variance composition

---

# Summary

BatchVaria provides a coherent and extensible framework for evaluating batch correction in high-dimensional expression data. By combining variance decomposition, structured tracking of results across multiple assays, and intuitive summary outputs, the package enables users to quantitatively assess how different correction strategies impact both unwanted technical variation and biological signal. The integration of variance composition tables, relative change metrics, and visual summaries supports transparent comparison of correction methods within a single workflow. Together, these features facilitate reproducible and interpretable evaluation of batch correction, addressing a key gap in many transcriptomic analysis pipelines.

---
# Session Information
```{r session}
sessionInfo()
```

