---
title: "BatchVaria Quickstart: Evaluating Batch Correction Performance"
author:
- name: Nicholas Moir
  affiliation: University of Edinburgh
  email: nicholas.moir@ed.ac.uk
package: BatchVaria
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{BatchVaria Quickstart: Evaluating Batch Correction Performance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, message=FALSE, warning=FALSE}
library(BatchVaria)
library(SummarizedExperiment)
```

# Overview

High-throughput transcriptomic datasets frequently contain unwanted technical variation arising from batch effects. Numerous batch correction methods exist, but selecting an appropriate method is challenging because correction can remove biological signal or distort sample relationships. BatchVaria provides a framework for quantifying variance components, tracking variance changes after correction, and evaluating correction performance using multiple metrics, enabling evidence-based selection of batch correction strategies.
<br>
<br>
While tools such as ComBat, limma, and BatchQC provide batch correction and diagnostic visualisations, BatchVaria focuses specifically on quantitative evaluation of correction performance across multiple assays and correction strategies within a unified framework. BatchVaria unifies variance decomposition, tracking, and comparison into a single coherent workflow.

---

# Installation

```{r installation}
# Install from Bioconductor (devel)
# BiocManager::install("BatchVaria")
```

# Example Dataset

We begin with a synthetic dataset built to reflect the situation batch
correction is hardest in: batch and biology are *partially confounded*. Most
control samples sit in one batch and most treated samples in the other, so the
two effects are not cleanly separable.

```{r runCorrection}
## exampleBatchVaria() draws random data; seed for a reproducible vignette
set.seed(1)
bv <- exampleBatchVaria(confounding = 0.6)

table(
    batch = bv$batch,
    group = bv$group
)
```

Correction methods are registered rather than hard-coded, so the set available
is extensible in the same way the variance engines are:

```{r methods}
availableCorrectionMethods()
```

This vignette uses `combat` throughout, so that the comparison below isolates
the effect of the design rather than differences between methods.

We then apply ComBat twice. The first call corrects for batch alone. The second
additionally names `group` as a variable to preserve, so the batch effect is
estimated adjusted for the named preservation variable.

```{r corrections, message=FALSE}
bv <- runCorrection(
    bv,
    method = "combat", batch = "batch",
    newAssayName = "naive"
)

bv <- runCorrection(
    bv,
    method = "combat", batch = "batch", preserve = "group",
    newAssayName = "protected"
)

assayNames(bv)
```

`raw_noise` is a deliberately distorted assay supplied with the example data,
kept here as a reference point for what a genuinely destructive transformation
looks like. `raw_center` and `raw_scale` are no-op controls: centring and
scaling change no variance fraction, so a diagnostic that reports a change on
them is reporting an artefact of the diagnostic rather than of the data.

---

# Variance Profiling 

We explicitly quantify the contribution of known covariates.

```{r profile}
assaysUsed <- c("raw", "naive", "protected", "raw_noise")

bv <- profileVariance(
    bv,
    formula = ~ batch + group,
    assays = assaysUsed,
    methods = c("anova", "variancePartition")
)
```
This step fits the specified model to each assay and decomposes variance into
batch (unwanted), group (biological) and residual (unexplained). Two engines are
run: `anova`, which fits linear models per feature and partitions variance with
Type II sums of squares, and `variancePartition`, which fits mixed models. Every
result is stored in an internal variance ledger keyed on assay, method, formula
and term.

Because the ledger now holds two decompositions per assay, summaries must say
which one they mean. Asking for a table without choosing is an error rather than
a silent mixture, since the engines' terms are not interchangeable:

```{r ambiguous, error=TRUE}
varianceTable(bv, assays = assaysUsed)
```

---

# Variance Summary Tables

A core output of BatchVaria is a pair of summary tables.

```{r table}
vt <- varianceTable(bv, assays = assaysUsed, method = "anova")
```
---

# Variance Composition Table

This table shows the percentage of variance explained by each component for each assay.

```{r percent}
vt$percent
```
---

# Variance Change Relative to Raw

This table shows how each correction changed variance relative to the raw data.
Every entry is a change in a *fraction*, which constrains what it can be read
to mean; the next section sets out what that constraint is.

```{r delta}
varianceTable(
    bv,
    assays = assaysUsed,
    method = "anova",
    formatDelta = TRUE
)$delta
```
---

# Total Variance: Reading Fractions Correctly

Every number above is a *fraction*. Within an assay, the fractions sum to
one. This has a consequence that must be accounted for before any of these
tables can be interpreted: **removing variance associated with one term
mechanically inflates the fraction attributed to every remaining term**,
even when the absolute variance of those terms has not increased at all.

`assayVariance()` supplies the missing denominator.

```{r totalvar}
av <- assayVariance(bv, assays = assaysUsed)
av
```

Multiplying a fraction by its assay's total variance converts it to an
absolute quantity, which is directly comparable across assays.

```{r absolute}
vr <- varianceResults(bv, method = "anova")

absVar <- merge(
    vr[, c("assay", "term", "variance_fraction")],
    av[, c("assay", "total_variance")],
    by = "assay"
)
absVar$absolute <- round(absVar$variance_fraction * absVar$total_variance, 1)

## as.data.frame(), because pivot_wider() returns a tibble and tibble
## prints three significant figure.
as.data.frame(tidyr::pivot_wider(
    absVar[, c("assay", "term", "absolute")],
    names_from = "term",
    values_from = "absolute"
))
```

Read the `group` row across the assays: 144.6 in `raw`, 126.0 after the naive
correction and 127.0 after the protected one. In absolute terms group variance
barely moves. Its *fraction* meanwhile climbs from 11.3 to 15.2 per cent,
because the denominator fell from 1281 to 829. The fraction rose because
something else was removed, not because biological signal grew.

---

# Two Engines, One Answer -- Until the Design Stops Cooperating

BatchVaria harmonises independent variance-profiling engines onto a common
schema: the same columns, and the same names for the two terms that mean the
same thing in any engine -- `residual` and `shared`. The two runs above reach
their decompositions by different routes: `anova` fits a linear model per
feature and partitions with Type II sums of squares, while
`variancePartition` fits mixed models via `lme4`.

```{r concordance}
cmp <- varianceResults(bv, assayName = "raw")[
    , c("method", "term", "variance_fraction")
]

as.data.frame(
    tidyr::pivot_wider(
        cmp,
        names_from = "method",
        values_from = "variance_fraction"
    )
)
```

Here they disagree, and the disagreement is the point. On an orthogonal design
they do not:

```{r orthogonal, message=FALSE}
orth <- profileVariance(
    exampleBatchVaria(confounding = 0),
    ~ batch + group,
    assays = "raw",
    methods = c("anova", "variancePartition")
)

## Printed at full precision deliberately: at three significant figures
## both columns read 0.163 and the agreement below cannot be checked.
as.data.frame(
    tidyr::pivot_wider(
        varianceResults(orth, assayName = "raw")[
            , c("method", "term", "variance_fraction")
        ],
        names_from = "method",
        values_from = "variance_fraction"
    )
)
```

Every term the two engines share agrees to floating-point precision. `shared`
is exactly zero, because an orthogonal design leaves nothing unattributable,
and `variancePartition` reports `NA` for it because a mixed model has no such
parameter to report.

The divergence above is therefore a property of the design, not of either
implementation. The two engines estimate different quantities:
`variancePartition` reports how much variance is *associated with* a term,
`anova` with Type II sums of squares reports how much is *uniquely attributable
to* it. When a design is orthogonal these coincide, which is why the two tables
differ as they do. When batch and group are partially aliased they cannot coincide,
because roughly 11 per cent of the variance here is consistent with either term
and identifiable to neither.

`anova` reports that share separately, as `shared`. A per-gene mixed model has
no corresponding parameter - and could not, since the effects are not
separately identifiable in this design - so the ambiguous variance is
reflected in both terms, each rising from about 11 per cent to about 17.
Neither number is wrong for the quantity it estimates; they simply stop being
interchangeable once the design cannot separate the effects.

The gap between them is therefore a measurement of the design rather than a
disagreement between tools: it says how much of the variance the experiment
permits you to attribute at all. Running both, and seeing them coincide on a
clean design and separate on a compromised one, is more informative than
trusting either alone.

---

# Interpretation

The naive correction looks like a success. Read the fraction table on its own:
batch variance halves, from 11.0 per cent to 5.6, and group variance *rises*,
from 11.3 per cent to 15.2. Unwanted variation down, biological signal up. By
the standard before-and-after comparison - decompose variance, correct,
decompose again - this is a correction working exactly as intended.

It is not. Three things the fraction table cannot show say otherwise.

**Total variance fell by 35 per cent**, from 1281 to 829. The group *fraction*
rose because the denominator collapsed, not because biology was preserved. In
absolute terms group variance fell from 145 to 126. This is the compositional
artefact: within an assay the fractions must sum to one, so removing variance
from any term mechanically inflates the share of every other. A rising fraction
is not evidence of preserved signal, and reading it as such is the easiest way
to certify a bad correction as a good one.

**The batch effect was not removed.** 5.6 per cent remains, half of what was
there. The correction cost a third of the dataset's variance and did not finish
the job it was paying for.

**`shared` went negative**, from +11.4 per cent to -3.5. The correction did not
merely fail to resolve the confounding; it distorted the design's
attributability, leaving the model terms explaining more jointly than they do
separately.

Protecting the covariate addresses the second of these. Telling ComBat that
`group` matters lets it estimate the batch effect adjusted for biology, and
batch variance drops to 1.1 per cent - essentially removed - while total
variance falls less far, to 927. It does not rescue biological signal on this
data: absolute group variance is 126 for the naive correction and 127 for the
protected one. The case for protecting the covariate here is that the correction
works, not that the biology survives better.

That balance is specific to this level of confounding. This vignette uses a
single value, `confounding = 0.6`; the trade-off moves with it, because as
batch and group approach aliasing, protecting the covariate increasingly means
protecting the batch effect along with it. Raising or lowering the parameter
will change which correction looks better on absolute group variance, so treat
that comparison as a property of the design rather than a general ranking of
the two calls.

The sign of `shared` is the more dependable signal. Across confounding levels
it is consistently negative after naive correction and positive after protected
correction, so a correction that drives it below zero has distorted the
design's attributability regardless of what the other fractions report. Being a
sign rather than a magnitude, that reading needs no threshold and no reference
distribution - but note it is demonstrated here on simulated data with one
correction method and one design shape.

For contrast, `raw_noise` shows what a destructive transformation looks like
under the same instruments. Its batch and group fractions both fall, which reads
as damage. But its total variance *more than doubles*, to 3077, so in absolute
terms every component grew: nothing was removed, the batch effect was diluted.
`basisRetention()` is unambiguous about the difference.

```{r retention}
basisRetention(bv, assays = assaysUsed)
```

It retains 0.37 of the original variance structure against 0.99 for both ComBat
runs, because added noise occupies directions the raw data did not. Note what
this measure does and does not separate: it distinguishes a correction from a
distortion, not a good correction from a bad one. Both ComBat runs stay on the
original axes because ComBat is a per-feature linear shift.

The general rule: a variance fraction is a statement about *composition*, not
about *quantity*. Use fractions to compare terms within an assay, and totals or
absolute variances to compare the same term across assays.

This is also the answer to what BatchVaria offers over calling the underlying
tools twice. Decomposing variance before and after correction is exactly what
certified the naive correction as successful. Recording each decomposition
alongside the total variance it is relative to, the share the design cannot
attribute, and the structure the correction preserved is what catches it.

---

# Summary Figure

`evaluateCorrections()` runs the diagnostics above in one call and returns them
in a form the plotting layer reads directly.

```{r summaryfig, fig.width=10, fig.height=6, out.width="100%"}
ev <- evaluateCorrections(
    bv,
    baseline = "raw",
    assays = assaysUsed,
    diagnostics = "variance"
)

plotVarianceDelta(ev)
```

One group of bars per model term, one bar per assay, and the height is the
signed change in that term's fraction relative to `raw`.  Read left to right: both corrections cut `batch`, the protected run cutting it roughly twice as hard. Both raise `residual`. The difference that matters is
`shared` - the naive run drives it down by 0.149 while the protected run moves
it by 0.044, which is the confounded biology being converted into apparent
residual rather than removed. `raw_noise` cuts `batch` too, by dilution rather
than correction, which is why no single bar here is sufficient on its own.

# Why This Matters

Batch correction is frequently applied as a routine preprocessing step, yet it is often performed without systematic or quantitative validation. This can lead to suboptimal outcomes, where overcorrection removes meaningful biological signal or undercorrection leaves residual technical artefacts that confound downstream analyses. BatchVaria addresses this gap by providing a transparent and reproducible framework for evaluating correction strategies. By enabling direct comparison of variance components across multiple corrected assays, the package supports quantitative validation of correction outcomes and facilitates more robust, interpretable analytical pipelines.

---

# BatchVaria Workflow Summary

The typical BatchVaria workflow is:

1. Load expression data into BatchVariaData
2. Apply batch correction methods with runCorrection(), naming any
   variables to preserve
3. Profile variance using profileVariance()
4. Summarise variance using varianceTable(method = ...), naming the engine
5. Convert fractions to absolute variance using assayVariance()
6. Compare corrections using delta variance and basisRetention()
7. Visualise variance composition

---

# Summary

BatchVaria provides a coherent and extensible framework for evaluating batch correction in high-dimensional expression data. By combining variance decomposition, structured tracking of results across multiple assays, and intuitive summary outputs, the package enables users to quantitatively assess how different correction strategies impact both unwanted technical variation and biological signal. The integration of variance composition tables, relative change metrics, and visual summaries supports transparent comparison of correction methods within a single workflow. Together, these features facilitate reproducible and interpretable evaluation of batch correction, addressing a key gap in many transcriptomic analysis pipelines.

---
# Session Information
```{r session}
sessionInfo()
```

