---
title: "BatchVaria Quickstart: Evaluating Batch Correction Performance"
author:
- name: Nicholas Moir
  affiliation: University of Edinburgh
  email: nicholas.moir@ed.ac.uk
package: BatchVaria
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{BatchVaria Quickstart: Evaluating Batch Correction Performance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, message=FALSE, warning=FALSE}
library(BatchVaria)
library(SummarizedExperiment)
```

# Overview

High-throughput transcriptomic datasets frequently contain unwanted technical variation arising from batch effects. Numerous batch correction methods exist, but selecting an appropriate method is challenging because correction can remove biological signal or distort sample relationships. BatchVaria provides a framework for quantifying variance components, tracking variance changes after correction, and evaluating correction performance using multiple metrics, enabling evidence-based selection of batch correction strategies.
<br>
<br>
While tools such as ComBat, limma, and BatchQC provide batch correction and diagnostic visualisations, BatchVaria focuses specifically on quantitative evaluation of correction performance across multiple assays and correction strategies within a unified framework. BatchVaria unifies variance decomposition, tracking, and comparison into a single coherent workflow. 
---

# Installation

```{r installation}
# Install from Bioconductor (devel)
# BiocManager::install("BatchVaria")
```

# Example Dataset

We begin with a synthetic dataset designed to reflect realistic batch correction scenarios. The example dataset contains simulated gene expression data with both biological group effects and batch effects. Several pre-generated corrected assays are included to illustrate different correction outcomes, including over-correction, under-correction and noise distortion. A ComBat correction is also performed as part of the workflow.


```{r runCorrection}
## exampleBatchVaria() draws random data; seed for a reproducible vignette
set.seed(1)
bv <- exampleBatchVaria()

bv <- runCorrection(bv, method = "combat", batch = "batch", assayName = "raw")

bv
```

```{r display_assays}
assayNames(bv)
```

There are now 5 assays contained within bv, the BatchVariaData class instance. The "raw" assay is uncorrected data, "raw_combat" is ComBat corrected, "raw_center" is mean-centred data, "raw_scale" is scaled data and "raw_noise" is a distorted correction.

---

# Variance Profiling 

We explicitly quantify the contribution of known covariates.

```{r profile}
assaysUsed <- c("raw", "raw_combat", "raw_center", "raw_scale", "raw_noise")

bv <- profileVariance(
    bv,
    formula = ~ batch + group,
    assays = assaysUsed,
    methods = c("variancePartition")
)
```
This step fits the specified model to each assay and decomposes variance into batch (unwanted), group (biological) and residual (unexplained). Results are stored within an internal variance ledger.

---

# Variance Summary Tables

A core output of BatchVaria is a pair of summary tables.

```{r table}
vt <- varianceTable(bv, assays = assaysUsed)
```
---

# Variance Composition Table

This table shows the percentage of variance explained by each component for each assay.

```{r percent}
vt$percent
```
---

# Variance Change Relative to Raw

This table shows how each correction changed variance relative to the raw data.

```{r delta}
varianceTable(bv, assays = assaysUsed, formatDelta = TRUE)$delta
```
---

# Total Variance: Reading Fractions Correctly

Every number above is a *fraction*. Within an assay, the fractions sum to
one. This has a consequence that must be accounted for before any of these
tables can be interpreted: **removing variance associated with one term
mechanically inflates the fraction attributed to every remaining term**,
even when the absolute variance of those terms has not increased at all.

`assayVariance()` supplies the missing denominator.

```{r totalvar}
av <- assayVariance(bv, assays = assaysUsed)
av
```

Multiplying a fraction by its assay's total variance converts it to an
absolute quantity, which is directly comparable across assays.

```{r absolute}
vr <- varianceResults(bv, method = "variancePartition")

absVar <- merge(
    vr[, c("assay", "term", "variance_fraction")],
    av[, c("assay", "total_variance")],
    by = "assay"
)
absVar$absolute <- absVar$variance_fraction * absVar$total_variance

tidyr::pivot_wider(
    absVar[, c("assay", "term", "absolute")],
    names_from = "term",
    values_from = "absolute"
)
```

---

# Interpretation

An effective correction should reduce variance attributable to unwanted
technical factors while preserving variance associated with the biological
variables of interest. Judging whether it has done so requires reading the
fraction table and the total variance together, because three of the four
transformations in this example are misread if the fractions are taken at
face value.

**ComBat.** The batch fraction falls sharply, from 16.3% to 2.2%, and the
group fraction *rises* from 16.4% to 18.4%. Read alone, that looks like
unwanted variation removed and biological signal preserved or even
enhanced. It is not. Total variance falls from 1225 to 902, so in absolute
terms the group component drops from 201 to 166 — ComBat removed roughly
**18% of the biological signal** along with the batch effect. The group
fraction rose only because the denominator shrank faster than the
numerator. This is the correction performing reasonably well overall, but
the cost to biology is invisible in the fraction table and plainly visible
once the denominator is shown.

**Scaling.** The composition is unchanged to the decimal — 16.3% / 16.4% /
67.3%, identical to raw. The fraction table therefore reports scaling as a
no-op. Total variance tells a different story: it falls from 1225 to 992,
because multiplying an assay by a constant scales every variance component
equally. Nothing was corrected, but the data were materially altered, and
fractions alone cannot see that.

**Added noise.** Batch and group fractions both fall (to 9.6% and 10.5%),
which reads as a correction that damaged the biology while partially
addressing batch. In absolute terms both components *increased* — batch
from 199 to 291 and group from 201 to 320 — because noise adds variance
everywhere. Nothing was removed at all; the batch effect was diluted, not
corrected. Only the residual fraction, and the near-tripling of total
variance to 3036, reveal what actually happened.

**Mean-centring** is the one case the fractions get right: composition and
total variance are both unchanged, because per-feature centring shifts
means without altering per-feature variances.

The general rule: a variance fraction is a statement about *composition*,
not about *quantity*. Use fractions to compare terms within an assay, and
absolute variances to compare the same term across assays. Reporting a
correction as signal-preserving on the basis of a rising fraction is the
most common way to reach a wrong conclusion from a right number.

# Summary Figure
```{r fig.width=10, fig.height=7, out.width="100%"}
plotVarianceComposition(bv, method = "variancePartition")
```

# Why This Matters

Batch correction is frequently applied as a routine preprocessing step, yet it is often performed without systematic or quantitative validation. This can lead to suboptimal outcomes, where overcorrection removes meaningful biological signal or undercorrection leaves residual technical artefacts that confound downstream analyses. BatchVaria addresses this gap by providing a transparent and reproducible framework for evaluating correction strategies. By enabling direct comparison of variance components across multiple corrected assays, the package supports quantitative validation of correction outcomes and facilitates more robust, interpretable analytical pipelines.

---

# BatchVaria Workflow Summary

The typical BatchVaria workflow is:

1. Load expression data into BatchVariaData
2. Apply batch correction methods 
3. Profile variance using profileVariance()
4. Summarise variance using varianceTable()
5. Compare corrections using delta variance
6. Visualise variance composition

---

# Summary

BatchVaria provides a coherent and extensible framework for evaluating batch correction in high-dimensional expression data. By combining variance decomposition, structured tracking of results across multiple assays, and intuitive summary outputs, the package enables users to quantitatively assess how different correction strategies impact both unwanted technical variation and biological signal. The integration of variance composition tables, relative change metrics, and visual summaries supports transparent comparison of correction methods within a single workflow. Together, these features facilitate reproducible and interpretable evaluation of batch correction, addressing a key gap in many transcriptomic analysis pipelines.

---
# Session Information
```{r session}
sessionInfo()
```

